---
title: "RNA-Seq Differential Expression Analysis Report"
{{SUBTITLE_LINE}}
author: "Bioinformatics Shared Resources at VCU"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: cosmo
    highlight-style: tango
    df-print: paged
    code-fold: true
    embed-resources: true
    # v1.4.0 fix (2026-05-04): full-page layout + widened content grid so
    # embedded DT tables (multi-column GSEA results) and side-by-side
    # heatmaps render without horizontal scrolling. Default body-width is
    # ~700-800px which truncates long Description columns.
    page-layout: full
    grid:
      body-width: 1400px
      sidebar-width: 250px
      margin-width: 200px
params:
  rds_path: ""
  analyst: "Mikail Bala"
  brs_ticket: ""
  genome_assembly: "primary assembly"
  sections_dir: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)

rds_data   <- readRDS(params$rds_path)
results    <- rds_data[[1]]
comparisons <- rds_data[[2]]
out_dirs   <- rds_data[[3]]
pca_plot   <- rds_data[[4]]
pca_plotly <- rds_data[[5]]
pca_3d     <- rds_data[[6]]
annotation <- rds_data[[7]]

# Phase 7.2: Tool versions from a single source of truth. Upstream tool
# versions live in `inst/extdata/upstream_versions.yml` (bump them there
# when nf-core/rnaseq is upgraded). R package versions are read live so
# they always match the renv-pinned lockfile.
upstream <- yaml::read_yaml(
  system.file("extdata/upstream_versions.yml", package = "bisrDE")
)

# Defensive version readers: `packageVersion()` fails if the package was
# loaded via `devtools::load_all()` (no DESCRIPTION on the install path).
# Fall back to reading DESCRIPTION via `system.file()` which IS supported
# under load_all, then to "dev" as a last resort.
.pkg_version <- function(pkg) {
  v <- tryCatch(as.character(utils::packageVersion(pkg)),
                error = function(e) NULL)
  if (!is.null(v)) return(v)
  desc_path <- tryCatch(system.file("DESCRIPTION", package = pkg),
                        error = function(e) "")
  if (nzchar(desc_path) && file.exists(desc_path)) {
    return(as.character(read.dcf(desc_path, fields = "Version")[1, 1]))
  }
  "dev"
}
r_versions <- list(
  deseq2          = .pkg_version("DESeq2"),
  edger           = .pkg_version("edgeR"),
  clusterprofiler = .pkg_version("clusterProfiler"),
  reactomepa      = .pkg_version("ReactomePA"),
  msigdbr         = .pkg_version("msigdbr"),
  bisrde          = .pkg_version("bisrDE")
)
```

## Overview

**Analyst:** `r params$analyst`

This report contains the results of differential expression analysis for `r length(comparisons)` comparisons.
It starts with an **Executive Summary** that distills each comparison into a one-paragraph
plain-English brief, then a **Sample Exploration using PCA** section that highlights sample
similarity and helps surface outliers. Next, we present the **Differential Expression Results**
section with full per-comparison subsections (volcano, heatmaps, top DEGs table, enrichment
plots). The report ends with manuscript-ready Methods text, references, and a glossary.
For further exploration of the results, refer to the output directories containing the raw
data files.

## Executive Summary

```{r exec-summary, results='asis'}
.fmt_pval <- function(x) if (is.na(x)) "n/a" else formatC(x, format = "e", digits = 2)
.top_n_genes <- function(df, n, direction) {
  if (is.null(df)) return(character(0))
  ok <- !is.na(df$padj) & df$padj < 0.05 &
    if (direction == "up") df$log2FoldChange >= 0.58
    else df$log2FoldChange <= -0.58
  if (!any(ok)) return(character(0))
  sub <- df[ok, , drop = FALSE]
  sub <- sub[order(sub$padj), , drop = FALSE]
  sub <- head(sub, n)
  if ("SYMBOL" %in% colnames(sub) && any(!is.na(sub$SYMBOL) & nzchar(sub$SYMBOL))) {
    out <- ifelse(is.na(sub$SYMBOL) | !nzchar(sub$SYMBOL),
                  as.character(sub$ENSEMBL_ID), sub$SYMBOL)
  } else {
    out <- as.character(sub$ENSEMBL_ID)
    if (length(out) == 0) out <- rownames(sub)
  }
  out
}
.top_pathway <- function(results_i) {
  best <- NULL
  for (slot in c("gsea", "kegg", "reactome", "hallmark")) {
    obj <- results_i[[slot]]
    if (is.null(obj)) next
    df <- as.data.frame(obj)
    if (nrow(df) == 0 || !"p.adjust" %in% colnames(df)) next
    top <- df[which.min(df$p.adjust), , drop = FALSE]
    label <- switch(slot, gsea = "GO", kegg = "KEGG",
                    reactome = "Reactome", hallmark = "Hallmark")
    cand <- list(backend = label, desc = top$Description,
                 padj = top$p.adjust)
    if (is.null(best) || cand$padj < best$padj) best <- cand
  }
  best
}

for (i in seq_along(comparisons)) {
  cmp <- comparisons[[i]]
  cat(sprintf("**%s** (%s vs %s).\n\n",
              cmp$name,
              paste(cmp$exp,  collapse = ", "),
              paste(cmp$ctrl, collapse = ", ")))

  if (is.null(results[[i]]) || is.null(results[[i]]$deseq)) {
    cat("- _No DE results available for this comparison._\n\n")
    next
  }

  df <- results[[i]]$deseq
  ok_up   <- !is.na(df$padj) & df$padj < 0.05 & df$log2FoldChange >=  0.58
  ok_down <- !is.na(df$padj) & df$padj < 0.05 & df$log2FoldChange <= -0.58
  n_up    <- sum(ok_up)
  n_down  <- sum(ok_down)
  total   <- n_up + n_down

  if (total == 0) {
    cat(paste0(
      "- **0 genes** passed FDR < 0.05 and |log2FC| ≥ 0.58. ",
      "Possible causes: small effect size, low replicate count, ",
      "or a contrast that doesn't capture the biology — ",
      "see Volcano + QC sections below.\n\n"
    ))
    next
  }

  top_up   <- .top_n_genes(df, 3, "up")
  top_down <- .top_n_genes(df, 3, "down")
  pathway  <- .top_pathway(results[[i]])

  cat(sprintf("- **%d** differentially expressed genes (%d up, %d down).\n",
              total, n_up, n_down))
  if (length(top_up) > 0) {
    cat(sprintf("- Top up-regulated: %s.\n",
                paste(sprintf("*%s*", top_up), collapse = ", ")))
  }
  if (length(top_down) > 0) {
    cat(sprintf("- Top down-regulated: %s.\n",
                paste(sprintf("*%s*", top_down), collapse = ", ")))
  }
  if (!is.null(pathway)) {
    cat(sprintf(
      "- Strongest enriched gene set: **%s** in %s (p.adjust = %s).\n",
      pathway$desc, pathway$backend, .fmt_pval(pathway$padj)
    ))
  } else {
    cat("- No enriched gene sets were detected by GO / KEGG / Reactome / Hallmark.\n")
  }
  cat("\n")
}
```

The full per-comparison breakdown (with volcano, heatmaps, DT tables, and dotplots) is in **Differential Expression Results** below.

## Pipeline

For this analysis we used the following steps:

1. **Data preprocessing**: Count data was filtered to remove genes with low expression.
2. **Normalization**: TMM normalization was applied using edgeR TMM function.
3. **Differential expression**: DESeq2 was used to identify differentially expressed genes.
4. **Functional analysis**: Gene Set Enrichment Analysis (GSEA) was performed using clusterProfiler.
5. **Thresholds**: Genes with adjusted p-value < 0.05 and |FC| >= 1.5 (equivalent to log2(FC) = 0.58) were considered differentially expressed.
6. **Genome annotation**: `r annotation`

As part of this pipeline we produce the following files for your downstream use:

```
output/
├── de_data/
│ ├── DESeq2_[comparison].csv
│ └── normalizedCounts_TMM[date].csv
├── gsea_data/
│ └── GO_Analysis_[comparison].csv
└── figures/
  ├── volcano/
  │   └── [comparison]_volcano.png
  ├── heatmap/
  │ └── [comparison]_heatmap_*.png
  ├── gsea/
  │ └── [comparison]_GSEA.png
  └── pca/
    ├── PCA_plot.png
    ├── allsamples_PCA_plot.html
    └── allsamples_PCA_plot3D.html
```

## Sample Exploration using PCA

PCA was performed to visualize the overall patterns of gene expression across samples and
to identify potential batch effects or outliers. **What we expect:** Samples with similar
expression profiles should cluster together, while dissimilar samples are expected to
separate into distinct clusters.

::: {.callout-tip title="How to read this"}
Each dot is a sample, projected onto the two (or three) axes (PC1, PC2, PC3) that capture the largest sources of variation across all genes. PCA is computed on TMM-normalized counts, so library-size differences are already accounted for. The percent-variance label on each axis tells you how much of the total spread that PC explains.
:::

::: {.callout-note title="What to look for"}
Replicates of the same `condition` should cluster together. PC1 typically captures the strongest biological signal (e.g. treatment vs control); PC2 often reflects a secondary structure (batch, time point, sex). Tight, well-separated clusters along PC1 are the best news; overlapping clusters across PC1 and PC2 may still be salvageable if your design has clear within-group consistency.
:::

::: {.callout-warning title="Common pitfalls"}
A sample that is a clear outlier on **only PC1** could just be deeper-sequenced — cross-check with the library-size barplot in the QC section before flagging it. **No separation** between conditions can mean low effect size, too few replicates, or batch confounding (samples from one prep batch dominate one PC). **Tight clustering of all samples** sometimes means the biology is subtle relative to between-replicate noise — check the per-sample log-CPM density curves for distribution mismatches.
:::

::: {.callout-tip title="Next steps"}
If a sample is an outlier across all three PCs, investigate (a) the samplesheet for sample swaps; (b) the lab notebook for failed library prep or contamination; (c) the correlation heatmap to see which other sample(s) it should look like. If conditions don't separate, consider whether your design captures the biology — time-course or dose-response studies often need more contrasts than the simple A vs B you are running here.
:::

### Static PCA (2D)

```{r}
#| fig-width: 6
#| fig-height: 4
print(pca_plot)
```

### Interactive PCA (2D)

*Hover a point for sample ID; drag to zoom; double-click to reset.*

```{r}
pca_plotly
```

### Interactive PCA (3D)

*Drag to rotate; scroll to zoom.*

```{r}
pca_3d
```

Standalone interactive PCA files are also produced for sharing or embedding elsewhere:

- 2D PCA: `r file.path(out_dirs$pca, "allsamples_PCA_plot.html")`
- 3D PCA: `r file.path(out_dirs$pca, "allsamples_PCA_plot3D.html")`

### Sample-level QC plots

These four sample-level QC plots complement the PCA above. Use them to spot outlier samples (poor depth, atypical distributions, mis-clustered replicates) before drawing conclusions from the differential-expression results.

::: {.callout-tip title="How to read these"}
Four complementary views of sample-level quality. The **correlation heatmap** asks "do replicates correlate?" using Spearman correlation on DE genes. The **vst distance heatmap** asks "do samples cluster by group?" using Euclidean distance on variance-stabilized counts. The **library-size barplot** asks "is sequencing depth uneven?" by summing raw counts per sample (top panel) and counting expressed genes per sample (bottom panel). The **hclust + density** asks "do all samples have similar expression distributions?" via a Ward dendrogram (left) and per-sample log-CPM density curves (right).
:::

::: {.callout-note title="What to look for"}
On the correlation heatmap, **dark red within-group blocks** (high Spearman ρ, ~0.95+) are healthy; cross-group ρ should be lower than within-group ρ. On the vst distance heatmap, **dark blue within-group blocks** are healthy (small distance = similar). The library-size barplot should show all samples within ~2-fold of each other and similar detected-gene counts. The hclust dendrogram should mostly cluster replicates; the density curves should overlap closely on the same axes.
:::

::: {.callout-warning title="Common pitfalls"}
**One discordant sample across multiple panels** is a strong outlier signal — much more reliable than a single panel disagreeing. **Library size differences > 3x** can drive PCA / clustering artefactually; consider down-sampling the deepest libraries before re-running. **Bimodal density curves** (two humps in log-CPM) often indicate failed rRNA depletion or 3'-bias. **Replicates in separate dendrogram clades** is the most concerning signal — confirm via the correlation heatmap before discussing with the wet lab.
:::

::: {.callout-tip title="Next steps"}
If you spot a clear outlier, the canonical workflow is: (1) drop it from the samplesheet, (2) re-run the pipeline, (3) compare DE results to confirm the outlier was suppressing real signal (or was a benign noise source). Don't drop more than one sample per group without consulting the experimental statistician — small N is the more common failure mode than outlier contamination. If everything looks healthy here, you can read the per-comparison DE results below with high confidence.
:::

**Sample × Sample correlation heatmap (Spearman, DE genes)**

```{r}
#| out-width: "900px"
#| fig-align: "center"
qc_corr_path <- file.path(out_dirs$qc, "qc_correlation_heatmap.png")
if (file.exists(qc_corr_path)) {
  knitr::include_graphics(qc_corr_path)
} else {
  cat("Correlation heatmap not available (no DE genes / <2 samples).")
}
```

**Sample × Sample Euclidean distance (vst-transformed counts)**

```{r}
#| out-width: "900px"
#| fig-align: "center"
qc_vst_path <- file.path(out_dirs$qc, "qc_vst_dist_heatmap.png")
if (file.exists(qc_vst_path)) {
  knitr::include_graphics(qc_vst_path)
} else {
  cat("vst distance heatmap not available.")
}
```

**Library size + detected genes per sample**

```{r}
#| out-width: "850px"
#| fig-align: "center"
qc_lib_path <- file.path(out_dirs$qc, "qc_libsize_detected.png")
if (file.exists(qc_lib_path)) {
  knitr::include_graphics(qc_lib_path)
} else {
  cat("Library size barplot not available.")
}
```

**Hierarchical clustering (Ward.D2) + per-sample log-CPM density**

```{r}
#| out-width: "1100px"
#| fig-align: "center"
qc_hcd_path <- file.path(out_dirs$qc, "qc_hclust_density.png")
if (file.exists(qc_hcd_path)) {
  knitr::include_graphics(qc_hcd_path)
} else {
  cat("Hierarchical clustering plot not available.")
}
```

```{r results-summary}
summary_table <- data.frame(
  Comparison    = character(),
  Experimental  = character(),
  Control       = character(),
  Total_DEGs    = integer(),
  Upregulated   = integer(),
  Downregulated = integer(),
  GO            = character(),
  KEGG          = character(),
  Reactome      = character(),
  Hallmark      = character(),
  stringsAsFactors = FALSE
)

for (i in seq_along(comparisons)) {
  if (!is.null(results[[i]])) {
    res_df <- results[[i]]$deseq
    if (!is.null(res_df)) {
      total_degs <- sum(!is.na(res_df$padj) & res_df$padj < 0.05 & abs(res_df$log2FoldChange) >= 0.58)
      up_degs    <- sum(!is.na(res_df$padj) & res_df$padj < 0.05 & res_df$log2FoldChange >= 0.58)
      down_degs  <- sum(!is.na(res_df$padj) & res_df$padj < 0.05 & res_df$log2FoldChange <= -0.58)

      yn <- function(x) ifelse(!is.null(x), "Yes", "No")

      summary_table <- rbind(summary_table, data.frame(
        Comparison    = comparisons[[i]]$name,
        Experimental  = paste(comparisons[[i]]$exp, collapse = ", "),
        Control       = paste(comparisons[[i]]$ctrl, collapse = ", "),
        Total_DEGs    = total_degs,
        Upregulated   = up_degs,
        Downregulated = down_degs,
        GO            = yn(results[[i]]$gsea),
        KEGG          = yn(results[[i]]$kegg),
        Reactome      = yn(results[[i]]$reactome),
        Hallmark      = yn(results[[i]]$hallmark),
        stringsAsFactors = FALSE
      ))
    }
  }
}
```

## Differential Expression Results

This section contains a high-level table summarizing the differential expression results,
followed by a subsection for each comparison that visualizes the results using a volcano
plot, heatmap, table of top hits, and GSEA results (shown as both a bubble plot and a table).

### Summary of Comparisons

Briefly, we report the following:

- Total number of comparisons analyzed: `r length(comparisons)`
- Comparison with the highest number of DEGs: `r if (nrow(summary_table) > 0) summary_table$Comparison[which.max(summary_table$Total_DEGs)] else "n/a"` (`r if (nrow(summary_table) > 0) max(summary_table$Total_DEGs) else 0` DEGs)
- Comparison with the lowest number of DEGs: `r if (nrow(summary_table) > 0) summary_table$Comparison[which.min(summary_table$Total_DEGs)] else "n/a"` (`r if (nrow(summary_table) > 0) min(summary_table$Total_DEGs) else 0` DEGs)

The table below summarizes differential expression results from all comparisons.
Specifically, the **Total_DEGs** column reports the number of genes with an adjusted
p-value < 0.05 and |FC| ≥ 1.5. The **Upregulated** and **Downregulated** columns break
down this count accordingly. The **GO**, **KEGG**, **Reactome**, and **Hallmark** columns
indicate whether each enrichment backend produced any results for that comparison
(`Yes`) or returned no enriched gene sets (`No`).

```{r display-summary-table}
kable(summary_table, caption = "")
```

```{r per-comparison, results='asis'}
sec_path <- file.path(params$sections_dir, "_comparison.qmd")
for (i in seq_along(comparisons)) {
  cat(knitr::knit_child(sec_path, envir = environment(), quiet = TRUE),
      sep = "\n")
}
```

## Utilizing IPA

The CSV Differential Expression output from DESeq2 (available in the results directory
provided alongside this report *./output/de_data/DESeq2_[comparison_name].csv*), can be
uploaded directly into **QIAGEN Ingenuity Pathway Analysis (IPA)** for self-exploration of
pathways predicted to be enriched by this experimental condition. Massey's BISR provides
access to VCU's license of IPA. If you do not already have an account associated with this
license, you may reach out to **morecockcm@vcu.edu** with your name, VCU health or VCU
email, and request for IPA. To perform a core expression analysis, login with your
credentials here: **https://analysis.ingenuity.com/pa** and follow the instructions [here](https://qiagen.my.salesforce-sites.com/KnowledgeBase/KnowledgeNavigatorPage?id=kA41i000000L6rMCAS).

We host an annual hands-on training for IPA at the beginning of the fall semester. Please
email BISR if you would like to be a part of this training. In the meantime, QIAGEN has a
playlist of user-friendly tutorials available on Youtube titled "QIAGEN IPA Training
Videos" the **Qiagen Digital Insights Youtube** page.

## Manuscript-Ready Text

### Methods

Raw RNA-Seq fastq files were processed by the VCU Massey Comprehensive Cancer Center Bioinformatics Shared Resource (BISR) using the NextFlow nf-core/rnaseq v`r upstream$nf_core_rnaseq$version` pipeline [4]. Briefly, this pipeline assesses sequencing quality using FastQC v`r upstream$fastqc$version` [1] before and after trimming, performs adaptor trimming with Trim Galore! v`r upstream$trim_galore$version` [5], and aligns sequencing reads to the `r params$genome_assembly` reference genome using STAR v`r upstream$star$version` [2] with transcriptome quantification by Salmon v`r upstream$salmon$version` [8]. Pipeline output includes gene expression raw count data and a comprehensive QC report compiled by MultiQC v`r upstream$multiqc$version` [3].

Differential expression analysis was performed using DESeq2 v`r r_versions$deseq2` [7]. Lowly expressed genes (fewer than three samples with ≥10 reads) were filtered out per DESeq2 methods [7] prior to normalization and differential expression testing. Significance was calculated using the Wald test and adjusted using the Benjamini-Hochberg False Discovery Rate (FDR). Significant differentially expressed genes (DEGs) are defined as those with an FDR < 0.05 and absolute fold-change ≥ 1.5 (log2 fold-change ≥ 0.58). Heatmaps were generated from edgeR v`r r_versions$edger` TMM-normalized counts [9].

Pre-ranked Gene Set Enrichment Analysis (GSEA) [10] was performed across all genes (regardless of significance), using log2 fold-change as the ranking metric, against four complementary gene set collections: (i) Gene Ontology terms (BP / MF / CC) via `clusterProfiler` v`r r_versions$clusterprofiler` [11]; (ii) KEGG pathways via the same package; (iii) Reactome pathways via `ReactomePA` v`r r_versions$reactomepa` [12]; and (iv) MSigDB Hallmark gene sets [6] retrieved via `msigdbr` v`r r_versions$msigdbr`. The orchestrating R package is `bisrDE` v`r r_versions$bisrde` (this report's source). All computational analyses were performed on VCU's High Performance Research Computing cluster.

### References

References are alphabetized by first-author surname. The bracketed numbers in the Methods text above resolve to entries below.

1) Andrews S. FastQC: A Quality Control Tool for High Throughput Sequence Data. Babraham Bioinformatics; 2010. <https://www.bioinformatics.babraham.ac.uk/projects/fastqc/>

2) Dobin A, Davis CA, Schlesinger F, Drenkow J, Zaleski C, Jha S, Batut P, Chaisson M, Gingeras TR. STAR: ultrafast universal RNA-seq aligner. Bioinformatics. 2013;29(1):15-21. doi:10.1093/bioinformatics/bts635

3) Ewels P, Magnusson M, Lundin S, Käller M. MultiQC: summarize analysis results for multiple tools and samples in a single report. Bioinformatics. 2016;32(19):3047-3048. doi:10.1093/bioinformatics/btw354

4) Ewels P, Peltzer A, Fillinger S, Patel H, Alneberg J, Wilm A, Garcia MU, Di Tommaso P, Nahnsen S. The nf-core framework for community-curated bioinformatics pipelines. Nat Biotechnol. 2020;38(3):276-278. doi:10.1038/s41587-020-0439-x

5) Krueger F. Trim Galore! v`r upstream$trim_galore$version`. 2023. <https://github.com/FelixKrueger/TrimGalore>

6) Liberzon A, Birger C, Thorvaldsdóttir H, Ghandi M, Mesirov JP, Tamayo P. The Molecular Signatures Database (MSigDB) hallmark gene set collection. Cell Syst. 2015;1(6):417-425. doi:10.1016/j.cels.2015.12.004

7) Love MI, Huber W, Anders S. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biol. 2014;15:550. doi:10.1186/s13059-014-0550-8

8) Patro R, Duggal G, Love MI, Irizarry RA, Kingsford C. Salmon provides fast and bias-aware quantification of transcript expression. Nat Methods. 2017;14(4):417-419. doi:10.1038/nmeth.4197

9) Robinson MD, McCarthy DJ, Smyth GK. edgeR: a Bioconductor package for differential expression analysis of digital gene expression data. Bioinformatics. 2010;26(1):139-140. doi:10.1093/bioinformatics/btp616

10) Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, Paulovich A, Pomeroy SL, Golub TR, Lander ES, Mesirov JP. Gene set enrichment analysis: A knowledge-based approach for interpreting genome-wide expression profiles. Proc Natl Acad Sci USA. 2005;102(43):15545-15550. doi:10.1073/pnas.0506580102

11) Yu G, Wang LG, Han Y, He QY. clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS. 2012;16(5):284-287. doi:10.1089/omi.2011.0118

12) Yu G, He QY. ReactomePA: an R/Bioconductor package for reactome pathway analysis and visualization. Mol Biosyst. 2016;12(2):477-479. doi:10.1039/c5mb00663e

### Required Acknowledgements

Please include the following statements in your acknowledgements manuscript section:

- "Services in support of the research project were provided by the VCU Massey Comprehensive Cancer Center Bioinformatics Shared Resource. Massey is supported, in part, with funding from NIH-NCI Cancer Center Support Grant P30 CA016059."

- "High Performance Computing resources provided by the High Performance Research Computing (HPRC) core facility at Virginia Commonwealth University (https://hprc.vcu.edu) were used for conducting the research reported in this work."

## Glossary

Plain-English definitions for the terms used throughout this report. Terms are alphabetized.

**baseMean.** The average normalized count across all samples for a given gene. Useful as a sanity-check: genes with `baseMean < 5` are typically too lowly-expressed to call DE reliably even when their `padj` is low.

**Benjamini-Hochberg (BH).** A method for adjusting p-values to control the False Discovery Rate when running many tests at once (one per gene, in our case). The result of BH adjustment is `padj`.

**contrast.** A single experimental comparison, e.g. `treated_vs_control`. Each contrast column in the samplesheet defines one contrast.

**DEG.** Differentially Expressed Gene — a gene that passes the project's significance thresholds (`padj < 0.05` AND `|log2FC| ≥ 0.58`).

**DESeq2.** The R package that fits a negative-binomial model per gene, estimates dispersion, and runs the Wald test that produces our DE results.

**edgeR.** A separate Bioconductor package; we use only its TMM normalization function (not its DE testing) to produce the count matrix that underlies heatmaps and PCA.

**FDR (False Discovery Rate).** The expected fraction of false positives among the genes you call "significant." Setting `padj < 0.05` means you accept ≤ 5% of your DEGs being noise on average. The whole point of multiple-testing correction.

**Gene Ratio.** On enrichment dotplots: `count of leading-edge genes in this set / total ranked genes`. Higher = more of the set's genes drove the enrichment signal.

**GO (Gene Ontology).** A controlled vocabulary describing gene function in three orthogonal axes: Biological Process (what the gene does), Molecular Function (how it does it), Cellular Component (where it does it).

**GSEA (Gene Set Enrichment Analysis).** A method for asking "are the genes in this set systematically up or down in my data?" without needing a pre-filtered DEG list. We run pre-ranked GSEA, ranking all genes by `log2FoldChange`.

**Hallmark.** MSigDB's curated set of 50 well-defined gene signatures representing high-level biological states (e.g. "EMT", "Hypoxia", "Inflammatory response"). Conservative; a Hallmark hit is usually robust.

**KEGG.** The Kyoto Encyclopedia of Genes and Genomes — a curated database of metabolic and signaling pathways. Smaller and more curated than Reactome.

**leading_edge.** On enrichment results: the subset of genes that drove the enrichment signal (i.e. before the running-sum statistic peaked). The "interesting" genes within an enriched set.

**log2FoldChange (log2FC).** Log2 of the ratio of mean expression between experimental and control groups. `+1` = 2x higher in experimental; `-1` = 2x lower; `0.58` ≈ 1.5-fold change (our project threshold).

**NES (Normalized Enrichment Score).** GSEA's enrichment score normalized for gene-set size, signed by direction (positive = enriched in up-regulated genes; negative = enriched in down-regulated). Comparable across gene sets, unlike the raw enrichment score.

**padj.** The Benjamini-Hochberg-adjusted p-value. **This** is the column to filter on for significance — not `pvalue`. `padj < 0.05` means you accept ≤ 5% FDR.

**PCA (Principal Component Analysis).** A dimensionality-reduction technique that projects the high-dimensional count data onto axes that capture the most variation. PC1 captures the most variance; PC2 the next-most orthogonal variance; etc.

**Reactome.** A peer-curated knowledgebase of biological pathways. Larger and more granular than KEGG; ~1500+ pathways vs. ~340 in KEGG.

**setSize.** On enrichment results: the number of genes in the gene set that overlap with your dataset's ranked list. Small (< 20) = very specific; large (> 200) = broad / generic.

**TMM (Trimmed Mean of M-values).** edgeR's normalization method. Adjusts for differences in library composition (not just library size); robust to a few highly-expressed genes dominating a sample.

**vst (variance-stabilizing transformation).** A DESeq2 transform that puts counts onto an approximately log-scale where variance is similar across the full expression range. Used here for the QC distance heatmap; **not** used for the DE testing itself (which works on raw counts).

**Wald test.** The hypothesis test DESeq2 uses to ask "is `log2FoldChange` significantly different from 0 for this gene?" Generates the raw `pvalue`; BH correction generates `padj`.

**z-score.** On the heatmaps: `(value - row_mean) / row_sd`, computed per gene. Centers + standardizes each gene's expression so a value of `+2` means "this sample is 2 standard deviations above the gene's average."
